\chapter{Colophon}
\label{chap:colophon}
\minitoc

Our look is the result of reader comments, our own experimentation, and feedback from distribution channels. Distinctive covers complement our distinctive approach to technical topics, breathing personality and life into potentially dry subjects.

The animals on the cover of \textit{Beginning Perl for Bioinformatics} are green frog (\textit{Rana clamitans}) and American bullfrog (\textit{Rana catesbeiana}) tadpoles.

Tadpoles are the larvae of frogs and toads. They are aquatic and when first hatched have large, round heads and long, flat tails. Through a complex process of metamorphosis, tadpoles change from small fishlike creatures to the more familiar frogs and toads. This process can take from 10 days to 3 years depending on the species.

During the first stages of metamorphosis, a tadpole's hind legs sprout, its head begins to flatten, and its tail becomes shorter. In its early life, a tadpole feeds primarily on diatoms, algae, and small quantities of zooplankton. As metamorphosis continues, it stops eating and begins to reabsorb its tail for sustenance while its digestive system changes from primarily vegetarian to carnivorous. During the final stages of metamorphosis, the tadpole's front legs appear, its jaws form, its skeleton hardens, and its gills disappear as the lungs develop. It soon begins to breathe air at the surface of the water. A short time later, the tadpole emerges from the water, reabsorbs the last of its tail, and hops off as a frog or a toad.

Mary Anne Weeks Mayo was the production editor and copyeditor for \textit{Beginning Perl for Bioinformatics}. Matt Hutchinson and Jane Ellin provided quality control. Edie Shapiro, Matt Hutchinson, and Derek DiMatteo provided production assistance. Ellen Troutman-Zaig wrote the index.

Ellie Volckhausen designed the cover of this book, based on a series design by Edie Freedman. The cover image is an original illustration created by Lorrie LeJeune. Emma Colby produced the cover layout with Quark\texttrademark XPress 4.1 using Adobe's ITC Garamond font.

Melanie Wang designed the interior layout, based on a series design by David Futato. Neil Walls converted the files from SGML to FrameMaker 5.5.6 using tools created by Mike Sierra. The text font is Linotype Birka; the heading font is Adobe Myriad Condensed; and the code font is LucasFont's TheSans Mono Condensed. The illustrations that appear in the book were produced by Robert Romano and Jessamyn Read using Macromedia FreeHand 9 and Adobe Photoshop 6. The tip and warning icons were drawn by Christopher Bing. This colophon was written by Lorrie LeJeune.
